---
title: "Programming Language Compensation Gap in the United States"
subtitle: "DA2/C2 - Term Project"
author: "Son Nam Nguyen"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
urlcolor: blue
fontsize: 9pt
geometry: margin=1.8cm
header-includes: |
  \usepackage{titling}
  \setlength{\droptitle}{-5em}
---

```{r chunks, echo=F}
#setting chunk options
knitr::opts_chunk$set(echo = F, message = F, warning = F, fig.align="center", fig.width = 6, fig.height = 4)
```

```{r setup}

# Loading packages with pacman
if (!require("pacman")) {
  install.packages("pacman")
}

pacman::p_load(tidyverse, modelsummary, kableExtra, fixest, estimatr, data.table, janitor, stringr)

#Reading csv from github repo
df <- read.csv("https://raw.githubusercontent.com/nszoni/da2-c1-term-project/main/stack-overflow-developer-survey-2021/survey_results.csv", header=TRUE, sep = ",", quote = "\"", fill = T, stringsAsFactors = F)

```

## Introduction

In this project, I will present my findings on the programming language compensation disparity among full-time developers in the United States. My aim is to show the expected differences in compensation depending on one knowing Python, SQL, Java, etc. controlled with employment, experience, education, and demographics confounders. My initial hypothesis is that languages in high demand like Python are worth learning to gain an advantage in the labor market.

For the analysis, I have used the [Stackoverflow Developer Survey](https://insights.stackoverflow.com/survey/2021) conducted in 2021. This report is based on a survey of **83,439 software developers from 181 countries around the world**. Respondents were recruited primarily through channels owned by Stack Overflow. The top sources of respondents were onsite messaging, blog posts, email lists, banner ads, and social media posts.

## Data Source

I consider the sample to be representative of the community. First, because it reaches out to potential respondents through an extensive number of channels. Yes, these mediums all assume that developers all have access to network connections, have their email, and so on, but it is extremely rare that people in the IT sector don't have those.

Looking back at earlier polls, the organization has made several changes throughout the years to reduce measurement errors in the variables. Instead of allowing respondents to offer arbitrary answers to questions like age, they grouped the variable into bins to eliminate age inconsistencies, at the cost of limiting the resolution with which age coefficients could be estimated. Another possible window for error could be recognized in the reported level of compensation. People **who believe that they are underpaid are less inclined to reveal their compensation, while those who are paid higher than average are eager to do so**. Also, I need to identify reported compensations that are not plausible given the employment, in other words, drop extreme values which are beyond a certain reliability threshold.

Like many others, the dataset lacks a variable that lets me determine the coding years spent with the programming languages. Therefore, it is unknown what share of their professional career was spent on working with the listed languages someone has worked with so far. Simply put, I can't estimate the coefficients for what should we expect in terms of remuneration when working an additional year with a language. Finally, **job titles/fields have always been a challenge to comprehend**. Titles frequently do not reflect what a professional does on a daily basis. Many job titles are not sufficiently differentiated from one another, and as a result, they tend to blur together (data people often call themselves developers). To put it another way, the sets of job tasks are rarely mutually exclusive to each other.

## Variables

Let me shortly clarify variables that are not that trivial what they represent. Each variable represent an entry in the survey which you can find [here](https://github.com/nszoni/da2-c1-term-project/blob/main/stack-overflow-developer-survey-2021/so_survey_2021.pdf). 

**Salaries are converted from user currencies to USD using the exchange rate on 2021-06-16**, and also converted to annual salaries assuming 12 working months and 50 working weeks. As a best practice, I will use the log of converted yearly compensations in thousand USD as my LHS variable.

Control variables I plan to focus on are **(1) Field of employment, (2) Size of organization, (3) Level of education (4) Gender, (5) Age, (6) Ethnicity (7) Number of known programming languages, (8) Years of coding**. I have also restricted the number of languages under the scope to the top five languages according to the survey which are **Javascript, HTML/CSS, Python, Java, SQL**.

## Data Munging

**To clean the raw data, I performed the following transformations:** (1) Dropped of columns not needed in the further analysis, (2) regrouped **Organization Size**, **DevType**, **Ethnicity** and **Gender** variables to decrease the number of factor levels, (3) cast the variables to their appropriate class structure, (4) fliped empty strings, "Prefer not to say" and its synonyms to NAs (5) filtered sample to observations working full-time in the United States.

In addition to the tidy-up, binary flags were generated for working in coding languages (JavaScript, HTML/CSS, Python, Java, SQL) along with the count of coding languages an individual has worked with and the log transformation of Compensation represented in thousand USD.


```{r dataclean}

#columns to work with
target <- c("EdLevel", "Country", "YearsCode", "DevType", "OrgSize", "LanguageHaveWorkedWith", "Age", "Gender", "ConvertedCompYearly", "US_State", "Employment", "Ethnicity")

df_clean <- df[ ,which((names(df) %in% target)==TRUE)]

#Data cleaning (casting, dimensionality aggregation, calculating flags for programming languages)
df_clean <- df_clean %>%
      mutate(
        #cast
        YearsCode = as.numeric(YearsCode),
        EdLevel = as.factor(EdLevel),
        Age = as.factor(Age),
        ConvertedCompYearly = as.numeric(ConvertedCompYearly),
        US_State = as.factor(US_State),
        Employment = as.factor(Employment),
        
        #log transformation
        ConvertedCompYearlyK = ConvertedCompYearly/1000,
        lConvertedCompYearlyK = log(ConvertedCompYearlyK),
        
        #regroup control variables
        CompSize = ifelse(OrgSize %in% c("Just me - I am a freelancer, sole proprietor, etc.", "10 to 19 employees", "2 to 9 employees"), "small", 
                          ifelse(OrgSize %in% c("20 to 99 employees","500 to 999 employees", "1,000 to 4,999 employees"), "medium", 
                                 ifelse(OrgSize %in% c("1,000 to 4,999 employees", "5,000 to 9,999 employees", "10,000 or more employees"), "large", NA))),
        Gender = as.factor(ifelse(grepl("Man", Gender, ignore.case=F), "Man", ifelse(grepl("Woman", Gender, ignore.case=TRUE), "Women", ifelse(grepl("Non-binary", Gender, ignore.case=TRUE), "Non-binary", NA)))),
        LanguageCount = as.integer(str_count(LanguageHaveWorkedWith, ";"))+1,
        Race = as.factor(ifelse(grepl("White", Ethnicity, ignore.case=TRUE), "White", "Colored")),
        Field = as.factor(ifelse(grepl("Dev", DevType), "Dev", 
                       ifelse(grepl("Data", DevType, ignore.case=TRUE), "Data", 
                              ifelse(grepl("Manager|Executive", DevType, ignore.case=TRUE), "Manager", 
                                     ifelse(grepl("site|admin", DevType, ignore.case=TRUE), "Sysadmin", "Other"))))),
        
        #create dummy for programming languages
        Python = c(0, 1)[1+grepl("Python", as.character(LanguageHaveWorkedWith))],
        SQL = c(0, 1)[1+grepl("SQL", as.character(LanguageHaveWorkedWith))],
        JavaScript = c(0, 1)[1+grepl("JavaScript", as.character(LanguageHaveWorkedWith))],
        HTML_CSS = c(0, 1)[1+grepl("HTML/CSS", as.character(LanguageHaveWorkedWith))],
        Java = c(0, 1)[1+grepl("Java", as.character(LanguageHaveWorkedWith))]
        )

#clean unknwon data
df_clean[df_clean == "Prefer not to say"] <- NA

#narrow down scope to full-time US developers
df_us <- df_clean %>% filter(Country == "United States of America" & Employment == "Employed full-time")

```

## Descriptive Statistics

The below descriptive table shows the key statistics of the numeric variables under the discussion. The distribution of **Years of coding is right-skewed with a mean of roughly 17 years**, while the dispersion is relatively high (measured at SD = 10.82). Starters have 1 year of coding, while veterans with the most experience reported 50 years of coding so far. 

Regarding yearly compensation, US developers follow an aggressively lognormal distribution in salaries, where the **mean is 265.4 thousand USD**, while the median is 125 thousand USD. 95% of the sample are paid less than 450 thousand USD, and **5% higher than that maxing out at a whopping 22 billion USD a year**. A minimum of 0 thousand USD wage means that there are observations are having a single-digit yearly compensation; calling for suspicion. According to [Minimum-wage.org](https://www.minimum-wage.org/federal), **the federal minimum yearly minimum wage in 2021 was $15,080.00/year**. Using that criterion, I've dropped rows with a yearly compensation below the minimum wage.

At last, **a developer working full-time in the US have experience in working with five or six programming languages on average**. The distribution of the variable is about bell-shaped, with a standard deviation of 2.9. According to the percentiles, 95% of the sample understands 11 or fewer programming languages, indicating that most two-digit numbers are uncommon in the developer community. **Some of the most experienced developers can code even in a total of 38 languages**.

```{r summary}

#define percentles and contruct summary table for numerics
P95 <- function(x){quantile(x,0.95,na.rm=T)}
P05 <- function(x){quantile(x,0.05,na.rm=T)}
datasummary( (`Years of coding` = YearsCode ) + 
             (`Yearly Compensation (K$)` = ConvertedCompYearlyK ) +
             (`Number of Languages Known` = LanguageCount) ~
             Mean + Median + SD + Min + Max + P05 + P95, 
             data = df_us,
             title = 'Descriptive statistics') %>% 
      kable_styling(latex_options = c("hold_position","striped")) %>% 
    add_footnote(c("Full-time developers in the United States"))

#filter out those who are earning below the federal minima wage
df_us <- df_us %>% filter(!(df_us$ConvertedCompYearly < 15000))
```

After clearing, filtering, and dropping implausible observations in the sample, I have explored once again the distribution of the years in coding, the number of languages a developer has worked with, and the distribution of programming language usage in the community. The graph in the [***Appendix***](#yearscoding) shows that the histogram of years of coding is still relatively right-skewed, with a lower mean of 16.78 years.

Then, I have illustrated the distribution of the number of coding languages known among developers in each employment field (see [***Appendix***](#nolanguages)). **Developers, Managers, and SysAdmins are having the highest average language known (with the larger IQRs), whereas people working in data have the lowest number of average languages they have worked with (low IQR)**. It is not surprising at all, considering that oftentimes managers are ex-developers, developers are more likely to work on both the front-end and back-end, while the toolkit of data people is usually reduced to the set of Python, R, and SQL. Additionally, the kernel density graph of all within field distributions can be characterized by a **left-peaked shape**.

**People still use Java and JavaSript the most** (more than 20% of the sample), as these languages have been in the discussion for decades and people have not yet fully transitioned to the now-booming ones like Python (sitting at roughly 16%). **SQL is also up there close to the leaders of the chart**, and personally, **I expect the SQL to resurface more among developers** because it is in high demand, especially in Data Engineering. The reason for that, new products following the Modern Data Stack mindset have to be SQL agnostic and thrive for supplying the community with a pipeline that can be managed by purely SQL (see [***Appendix***](#toplanguages)).

Before jumping into the set of models I have estimated, behind the scenes, I checked how numerical variables are correlated, and whether we have to address the possibility of multicollinearity (see [***Appendix***](#correlation)). Fortunately, **none of the pairs in my matrix had a correlation coefficient higher than 0.22**, meaning that we can reject the hypothesis that these are linear combinations of each other.

## Model

To present my estimated models, I chose to use an additive logic approach, in which I first estimated the unconditional link between compensation and (the five) programming languages, then afterward **added controlling factors relating to employment and demographics**. Variables are chosen on the basis that they are the key compensation mechanism variables in most studies. I could have included experience with a wide range of tools that were indicated in the survey and might be utilized to gain an advantage on the job market, but I refused to do so because it would have thrown off my coefficients/observations balance. The argument about the number of variables vs. the number of observations also relates to my decision not to estimate polynomial and interaction forms.

*Check full regression table in the [***Appendix***](#fulltable)*

```{r models}

#relevel factors (chose most observations as base to reduce SE)
df_us <- df_us %>% mutate(
  Field = relevel(Field, ref="Dev"),
  EdLevel = relevel(EdLevel, ref="Bachelor’s degree (B.A., B.S., B.Eng., etc.)"),
  Age = relevel(Age, ref="18-24 years old"),
  Race = relevel(Race, ref="White") 
)

#unconditional
reg1 <- feols(lConvertedCompYearlyK ~ Python + SQL + Java + JavaScript + HTML_CSS, data=df_us, vcov = "hetero")

#adding employment covariates
reg2 <- feols(lConvertedCompYearlyK ~ Python + SQL + Java + JavaScript + HTML_CSS + Field + CompSize, data = df_us, vcov = "hetero")

#adding demographics
reg3 <- feols(lConvertedCompYearlyK ~ Python + SQL + Java + JavaScript + HTML_CSS + Field + CompSize + EdLevel + Gender + Age + Race + YearsCode, data = df_us, vcov = "hetero")

#present results
msummary(list("Baseline" = reg1, "Employment" = reg2, "Demographics" = reg3),
         fmt="%.2f",
         gof_omit = 'DF|Deviance|Log.Lik.|F|R2 Adj.|AIC|BIC|R2 Within|R2 Pseudo',
         stars=c('*' = .05, '**' = .01),
         estimate = "{estimate} ({std.error}){stars}",
         statistic = NULL,
         coef_omit = c("Comp|EdLevel|Field|Gender|Age|YearsCode|Race"),
         title = "Coding Language Compensation Gap in the US",
         notes = c("Demographics: Adds (1) Age, (2) Gender, (3) Level of Education, (4) Years Coding, (5) White dummy", "Employment: Adds (1) Working Field, (2) Company Size", "Baseline: Estimates unconditionally")) %>% 
        column_spec(1:3, width = "9em") %>% 
        kable_classic(full_width = F, position = "center" , latex_options = "hold_position")

```


### Unconditional Estimates (N = 9,205)

The baseline model indicates that all of the languages except Python and JavaScript has a statistically significant impact on compensation levels, at 5%. **Those who know Java tend to have 8% higher yearly compensation in the US. Studying front-end skills like JavaScript and HTML/CSS are on average associated with 6 and 11% lower salaries than those who are not doing it, respectively, holding other covariates constant**. People knowing SQL also tend to earn less by 6% on average compared to those who are not familiar with it. The goodness-of-fit is relatively low with only 1% of the variance in y is explained by the variance of explanatory variables.

### Employment (Second-best model; N = 7,186)

Adding employment-specific variables like company size, and working department, we can observe that most of the explanatory variables have lost their significance on the 5% level, except HTML/CSS. The magnitude of coefficients all dropped or stayed the same for the regressors, however, the $R^2$ has also improved suggesting that included managed to cover a part of the variance in yearly compensation.

### Demographics (First-best model; N = 6,976)

Estimating my fully extended model where I included demographic control variables, the statistical significance of the explanatory variables changes once again. **Python now is significant and implies a 5% higher yearly compensation for those who have worked with it, all else being equal**. The coefficients of SQL, JavaScript, and HTML/CSS are still associated with lower compensation. **By working with SQL, compensation is expected to be lower on average by 4%, while lower by 8% in the case of HTML/CSS, assuming that other covariates are constant**. The $R^2$ of roughly 0.91 suggests that there are still a lot of unexplained variances in y, hence, the continuation of this project should incorporate other unknown variables.

## External Validity

To check the external validity of my best model which incorporates variation in employment and demographics. I have created additional subset of dataframes for the other four countries having the most observations in the survey dataset (UK and Ireland, France, India, and Germany)

Coefficents do not appear to be steady at all, which could be attributable to regional variances in skill demand. For example, **knowing Python and Java is substantially more favorable in India than in Germany (the former is related with 21% greater remuneration) as compared to the coefficients in the United States**. While Python is higher valued in India (16% greater salary), SQL knowledge is considered a no-go, with an average pay cut of 13%. Other notable discrepancies can be found in the estimated coefficients of **France, where workers with Python and JavaScript experience are penalized by 10% and 25% lower compensation, respectively, whereas Java developers are rewarded with a 21% higher salary**.

```{r robust}

coeff_intercept <- c()
coeff_python <- c()
coeff_sql <- c()
coeff_java <- c()
coeff_javascript <- c()
coeff_html_css <- c()
i=0

for (c in unique(df_clean[df_clean$Country=="India" | df_clean$Country=="Germany" | df_clean$Country=="United Kingdom of Great Britain and Northern Ireland" | df_clean$Country=="France", ]$Country)) {
  i=i+1
  reg <- feols(lConvertedCompYearlyK ~ Python + SQL + Java + JavaScript + HTML_CSS + Field + CompSize + EdLevel + Gender + Age + YearsCode + Race, data = df_clean[df_clean$Country==c, ], vcov = "hetero")
  coeff_intercept[c] <- reg$coefficients[[1]]
  coeff_python[c] <- reg$coefficients[[2]]
  coeff_sql[c] <- reg$coefficients[[3]]
  coeff_java[c] <- reg$coefficients[[4]]
  coeff_javascript[c] <- reg$coefficients[[5]]
  coeff_html_css[c] <- reg$coefficients[[6]]
}

ext <- data.frame(coeff_intercept, coeff_python, coeff_sql, coeff_java, coeff_javascript, coeff_html_css) 
rownames(ext) <- c("UK and Ireland","India","Germany","France")

ext %>% kbl(digits=2, caption = "Out-of-Sample Coeffcient Estimates", booktabs = T,
        col.names = c("Constant",
                      "Python",
                      "SQL",
                      "Java",
                      "JavaScript",
                      "HTML/CSS"))%>% 
  kable_styling(latex_options = c("hold_position","striped")) %>% 
  add_footnote(c("Dependent variable: Log Yearly Compensation in USD", "Model: Employment + Demographics regressors"))

```

## Summary

To sum up, **there isn't a good language to concentrate on if we want greater pay**. It undoubtedly depends on the country's job market, and preferences for specific languages shift regularly. **Python has a lot of potential, but it hasn't been widely adopted compared to languages that are on the slide in terms of popularity**. It was also surprising to find SQL in the top ten because it is a very basic language but that’s what makes it easier for a data team to migrate on new technologies. Please note that these are mere conditional expectations and we should not infer cause and effect from comparisons in observational data. 

## Conclusion

I've learned a lot about the present work market, and I believe I've addressed many of the major stumbling blocks in order to achieve my goal. In terms of constraints, I aggregated levels of factor variables to simplify my model, but I paid the price by being unable to see those coefficients at a finer level. There is still a lot of unexplained variation in compensation, which can be linked to the company's performance, seniority, the effort he or she puts in at work, and other types of compensation (e.g. shares, cafeteria, etc.). Beyond that, there are factors such as family background and social skills that can only help to reduce the variance. I'm looking forward to delving deeper into this link and using other sources to supplement or proxy the latent variables provided.

## Appendix {#appendix} 

### Distribution in years of coding {#yearscoding}

```{r distributions}

#distribution of years in coding
ggplot(aes(YearsCode), data = df_us) + 
  geom_histogram(colour="firebrick4", fill="firebrick1", bins = 20) +
  geom_vline(aes(xintercept=mean(YearsCode, na.rm=T)),
           color="firebrick3", linetype="dashed", size=1, alpha = 0.5) +
  geom_text(aes(mean(YearsCode, na.rm=T), 900, label=paste0("Mean: ",round(mean(YearsCode, na.rm=T), 2))), hjust = -0.3) +
  labs(title = 'Years of Coding among Full-time Developers in the US',
       caption = 'Source: Stackoverflow Developer Survey 2021',
       x = 'Years of Coding',
       y = 'Frequency') +
  theme_bw()

```

### Number of languages known {#nolanguages}

```{r}

#field distribution of number of known languages
df_us %>% filter(!is.na(Field)) %>% 
  ggplot(aes(x=Field, y=LanguageCount, fill=Field)) +
    geom_violin(width=1, alpha=0.5)+
    geom_boxplot(width=0.3) +
    labs(x = "",
         y = "No. Languages",
         subtitle = "Distributions across field types",
         title = "Total Coding Languages Known",
         caption = "Source: Stackoverflow Developer Survey 2021") +
  theme_bw() + 
  scale_fill_viridis_d(guide="none")

```

### Most popular languages distribution {#poplanguages}

```{r}

#share of top languages in teh community
df_us %>% dplyr::select(., Python:Java) %>% pivot_longer(cols = c("Python", "Java", "SQL", "JavaScript", "HTML_CSS")) %>% ggplot() +
  geom_col(aes(reorder(name, value/sum(value)), value/sum(value)), fill="firebrick3") +
  labs(title = 'Share of Top Languages among US Developers',
     caption = 'Source: Stackoverflow Developer Survey 2021',
     x = '',
     y = 'Share of Total (%)') +
  scale_y_continuous(labels=scales::percent) +
  theme_bw() +
  coord_flip()

```

### Correlation matrix {#correlation}

```{r scatter}

#checking correlations
df_us[,c("YearsCode", "ConvertedCompYearlyK", "LanguageCount")] %>% cor(use="complete.obs") %>% 
  kable(digits=3, caption = "Correlation Matrix", booktabs = T,) %>% 
  kable_styling(latex_options = c("hold_position","striped")) %>% 
  add_footnote(c("Pearson correlation coefficients"))

```

\pagebreak
### Full regression table {#fulltable}

```{r fulltable}

msummary(list("Baseline" = reg1, "Employment" = reg2, "Demographics" = reg3),
         fmt="%.2f",
         gof_omit = 'DF|Deviance|Log.Lik.|F|R2 Adj.|AIC|BIC|R2 Within|R2 Pseudo',
         stars=c('*' = .05, '**' = .01),
         estimate = "{estimate} ({std.error}){stars}",
         statistic = NULL,
         coef_rename = c("FieldData" = "Data",
                         "FieldManager" = "Manager",
                         "FieldOther" = "Other Field",
                         "FieldSysadmin" = "SysAdmin",
                         "CompSizemedium" = "Medium Company",
                         "CompSizesmall" = "Small Company",
                         "EdLevelAssociate degree (A.A., A.S., etc.)" = "Associate degree",
                         "EdLevelMaster’s degree (M.A., M.S., M.Eng., MBA, etc.)" = "Master's degree",
                         "EdLevelOther doctoral degree (Ph.D., Ed.D., etc.)" = "Doctoral degree",
                         "EdLevelPrimary/elementary school" = "Elementary School",
                         "EdLevelProfessional degree (JD, MD, etc.)" = "Professional degree",
                         "EdLevelSecondary school (e.g. American high school, German Realschule or Gymnasium, etc.)" = "Secondary School",
                         "EdLevelSome college/university study without earning a degree" = "University w/out degree",
                         "EdLevelSomething else" = "Other Education Level",
                         "GenderNon-binary" = "Non-binary",
                         "GenderWomen" = "Women",
                         "Age25-34 years old" = "25-34",
                         "Age35-44 years old" = "35-44",
                         "Age45-54 years old" = "45-54",
                         "Age55-64 years old" = "55-64",
                         "Age65 years or older" = "65+",
                         "AgeUnder 18 years old" = "18-",
                         "YearsCode" = "Years Coded",
                         "RaceColored" = "People of color"
                         ),
         title = "Full table: Coding Language Compensation Gap in the US") %>% 
        column_spec(1:3, width = "9em") %>% 
        kable_classic(full_width = F, position = "center" , latex_options = "hold_position")

```